Follow-up Samples

### 2017-04-24 submission/manuscript.Rmd

**Results.** There were large changes to the bacterial communities associated with treatment across the three groups. The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value=`r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value > 0.05). Although treatment brought about large intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Because the distribution of OTUs across patients and diagnosis groups was patchy, we used the Random Forest machine learning algorithm to identify groups of OTUs that allowed us to successfully distinguish between pre- and post-treatment samples for each of the diagnosis groups. However, across the three models, there was little overlap between the OTUs that were indicative of treatment. Next, we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas to determine whether individuals who underwent a treatment were more likely to have OTUs associated with normal colons. We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses. Although we could clearly differentiate pre- and post-treatment communities from the three diagnosis groups, only those patients that initially had carcinomas experienced a significant decrease in positive probability of having a carcinoma (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05). Finally, we tested whether the type of treatment impacted the microbiota of those diagnosed with carcinomas and were unable to detect any significant differences in characteristics of these communities between individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

@@

Colorectal cancer (CRC) is the third most common cause of cancer deaths in the United States [@jemal_cancer_2010; @haggar_colorectal_2009]. Disease mortality has significantly decreased, thanks to improvements in screening [@jemal_cancer_2010]. Despite this improvement, there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009]. Current estimates indicate that 20-30% of those who undergo treatment will experience recurrence and 30-50% of those patients will die [@hellinger_reoperation_2006; @ryuk_predictive_2014]. Identification of methods to assess patients' risk of recurrence is of great importance to reduce mortality and healthcare costs.

There is growing evidence that the gut microbiota is involved in the progression of CRC. Mouse-based studies have identified populations of *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* that alter disease progression [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009]. Furthermore, studies that alter the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces has shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014]. Collectively, these studies hypothesize that the microbiota is altering the amount of inflammation in the colon and with it the rate of tumorigenesis [@flynn_metabolic_2016].

Building on this evidence, several human-based studies have identified unique signatures of colonic lesions [@wang_structural_2012; @chen_decreased_2013; @chen_human_2012; @shen_molecular_2010; @kostic_genomic_2012; @feng_gut_2015]. One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012]. Others have proposed feces-based biomarkers that could be used to diagnose the presence of colonic adenomas and carcinomas [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014]. These studies have associated *Fusobacterium nucleatum* and other oral pathogens with lesions (adenoma, advanced adenoma, and carcinoma). They have also noted that the loss of bacteria generally thought to produce short chain fatty acids, which can suppress inflammation, is associated with colonic lesions. Gut bacteria have a role in tumorigenesis and there is promise that these populations may be useful biomarkers for aiding in the early detection of disease.

@@

We answered these questions by sequencing the V4 region of 16S rRNA genes amplified from fecal samples of individuals with adenoma, advanced adenoma, and carcinomas pre- and post-treatment. We used both classical community analysis to compare the alpha and beta-diversity of communities pre- and post treatment. Next, we generated Random Forest models to identify bacterial populations that were indicative of treatment for each diagnosis group. Finally, we measured the predictive probability of diagnosis models to determine whether treatment would alter the probability of being diagnosed as having a normal colon or the colon of the original diagnosis group. We found that treatment alters the composition of the gut microbiota; however, we only observed a shift from diseased to normal colon for those initially diagnosed with carcinomas. Understanding how the community responds to these treatments could be a valuable tool for identifying biomarkers to quantify the risk of recurrence.

@@

***Treatment alters the bacterial community structure of patients diagnosed with colonic lesions.***
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples [Table 1]. The structure of the microbial communities of the pre- and post-treatment samples differed [Figure 1A]. We found that the communities obtained pre- and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value = `r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`). There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05). Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D]. We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value`r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value`r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value`r beta_summary["Carcinoma", "PERMANOVA"]`). Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value > 0.05) [Table S1].

***The effects of treatment are not consistent across treatment groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups [Figure S1]. It is likely that high inter-individual variation and the patchy distribution of OTUs across individuals limited the statistical power of the test. To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. Because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that that models would overfit the data. Despite this restriction, the models performed well (adenoma AUC=`r format(min(adn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_treat_AUC_summary$ROC), digits = 2)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2)`). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. These results support the earlier community-wide analysis where we observed that the treatment had an impact on the overall community structure; however, the effect of treatment was not consistent across patients and diagnosis groups.

***Post-treatment samples from patients with carcinoma change towards a microbiota associated with normal***
Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons. To test this, we used an expanded cohort of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma [Table 2]. We then constructed Random Forest models to classify the pre and post-treatment samples as having their original diagnosis or having a normal colon. The models performed well (adenoma AUC=`r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)`; Figure S2). The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A]. Among the OTUs that were shared across the three models were those populations commonly considered as commensals (e.g. *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B]. Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [Figure S3]. Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model. A decrease in the positive probability would indicate that the colon more closely resembled that of a patient with a normal colon. There was no significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 4]. The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased. In fact, only `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) that were initially diagnosed with a carcinoma had a higher positive probability after treatment. One of those was re-diagnosed with carcinoma on the follow up visit. These results indicate that although there are changes in the microbiota associated with treatment, those experienced by the patients in the carcinoma group were the only ones where the change was directed towards what would be found in a normal colon

***Difficult to identify effects of specific treatments on the change in the microbiota.***
The type of treatment that the patients received varied across diagnosis groups. Those with adenomas and advanced adenomas received surgical resection (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "Y")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "Y")))`) or polyp removal during colonoscopy (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "N")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "N")))`) and those with carcinomas received surgical resection (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "no")))`), surgical resection with chemotherapy (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "no")))`), and surgical resection with chemotherapy and radiation (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "yes")))`). We focused on the patients with carcinoma and pooled those patients that received chemotherapy with those that received chemotherapy and radiation to improve our statistical power. We did not observe a significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P= `r chemo_rad_summary["dist", "bh"]`). Finally, the change in the positive probability was not significantly different between the two treatment groups (P=`r chemo_rad_summary["red_crc", "bh"]`). Due to the relatively small number of samples in each treatment group, it is difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota

@@

Understanding the effect of treatment on the microbiota of those initially diagnosed with carcinomas may have important implications into understanding disease recurrence. The microbiota changes associated with treatment among patients initially diagnosed with carcinomas resulted in community structures that resembled patients with normal colons. This would suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at removing bacteria associated with them. Interestingly, 6 of the 26 patients diagnosed with carcinomas had an elevated probability of having carcinomas after treatment. This leads us to hypothesize that these individuals may be at a high risk of recurrence. Interestingly, the `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30% [@hellinger_reoperation_2006; @ryuk_predictive_2014]), which would support our hypothesis . Unfortunately, because of the design of our study, it was not possible to determine whether individuals who had an increased positive probability for carcinoma experienced recurrence. Regardless, it is intriguing to think of using biomarkers from the microbiota to not only predict the presence of lesions, but to also assess the risk of developing them.

@@

***16S rRNA Gene Sequencing.***
Sequencing was completed as described by Kozich, et al. [@kozich_development_2013]. DNA extraction used the 96-well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf). The V4 variable region was amplified and the resulting product was split between three sequencing runs with normal, adenoma, and carcinoma evenly represented on each run. Each group was randomly assigned to avoid biases based on sample collection location. The pre- and post-treatment samples were sequenced on the same run.

@@

***Statistical Analysis.***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis. Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`). Comparisons between probabilities as well as overall OTU differences between pre- and post-treatment samples utilized a paired Wilcoxon ranked sum test. Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant. Unless otherwise stated the P-values reported are those that were BH corrected. Model rank importance was determined by obtaining the median MDA from the 100, 20 repeated 10-fold cross validation and then ranking from largest to smallest MDA.

@@

**Figure 1: General Differences between Adenoma, Advanced Adenoma, and Carcinoma Groups After Treatment.** A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`). Solid black points represent the median value for each diagnosis group. B) NMDS of the pre- and post-treatment samples for the adenoma group. C) NMDS of the pre- and post-treatment samples for the advanced adenoma group. D) NMDS of the pre- and post-treatment samples for the carcinoma group.

@@

**Table 1: Demographic Data of Pre and Post Treatment Cohort**

>>>>

**Results.** There were large changes to the bacterial communities associated with treatment across the three groups. The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`); however, there was not a significant difference between those with advanced adenoma and those with adenoma or carcinoma (P-value > 0.05). Although treatment brought about large intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Because the distribution of OTUs across patients and diagnosis groups was patchy, we used the Random Forest machine learning algorithm to identify groups of OTUs that allowed us to successfully distinguish between pre and post-treatment samples for each of the diagnosis groups. Although the three models successfully differentiated between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of treatment. Next, we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas to determine whether individuals who underwent a treatment were more likely to have OTUs associated with normal colons. We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses. Only those patients that initially had carcinomas experienced a significant decrease in positive probability of having a lesion (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05). Finally, we tested whether the type of treatment impacted the microbiota of those diagnosed with carcinomas and were unable to detect any significant differences in characteristics of these communities between individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

@@

Colorectal cancer (CRC) is the third most common cause of cancer deaths in the United States [@jemal_cancer_2010; @haggar_colorectal_2009]. Disease mortality has significantly decreased, thanks to improvements in screening [@jemal_cancer_2010]. Despite this improvement, there are still approximately 50,000 CRC-related deaths per year [@haggar_colorectal_2009]. Current estimates indicate that 20-30% of those who undergo treatment will experience recurrence and 30-50% of those patients will die [@hellinger_reoperation_2006; @ryuk_predictive_2014]. Identification of methods to assess patients' risk of recurrence is of great importance to reduce mortality and healthcare costs.

There is growing evidence that the gut microbiota is involved in the progression of CRC. Mouse-based studies have identified populations of *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* that alter disease progression [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009]. Furthermore, studies that alter the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces has shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014]. Collectively, these studies support the hypothesis that the microbiota can alter the amount of inflammation in the colon and with it the rate of tumorigenesis [@flynn_metabolic_2016].

Building upon this evidence, several human-based studies have identified unique signatures of colonic lesions [@wang_structural_2012; @chen_decreased_2013; @chen_human_2012; @shen_molecular_2010; @kostic_genomic_2012; @feng_gut_2015]. One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012]. Others have proposed feces-based biomarkers that could be used to diagnose the presence of colonic adenomas and carcinomas [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014]. These studies have associated *Fusobacterium nucleatum* and other oral pathogens with colonic lesions (adenoma, advanced adenoma, and carcinoma). They have also noted that the loss of bacteria generally thought to produce short chain fatty acids, which can suppress inflammation, is associated with colonic lesions. Gut bacteria have a role in tumorigenesis and there is promise that these populations may be useful biomarkers for aiding in the early detection of disease.

@@

We answered these questions by sequencing the V4 region of 16S rRNA genes amplified from fecal samples of individuals with adenoma, advanced adenoma, and carcinomas pre and post-treatment. We used both classical community analysis to compare the alpha and beta-diversity of communities pre and post treatment. Next, we generated Random Forest models to identify bacterial populations that were indicative of treatment for each diagnosis group. Finally, we measured the predictive probability of diagnosis models to determine whether treatment would alter the probability of being diagnosed as having a normal colon or the colon of the original diagnosis group. We found that treatment alters the composition of the gut microbiota; however, we only observed a shift from diseased to normal colon for those initially diagnosed with carcinomas. Understanding how the community responds to these treatments could be a valuable tool for identifying biomarkers to quantify the risk of recurrence.

@@

***Treatment alters the bacterial community structure of patients diagnosed with colonic lesions.***
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1]. The structure of the microbial communities of the pre and post-treatment samples differed [Figure 1A]. We found that the communities obtained pre and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value = `r format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = TRUE, digits = 2, nsmall = 2)`). There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05). Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D]. We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value=`r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value=`r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value=`r beta_summary["Carcinoma", "PERMANOVA"]`). Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value > 0.05) [Table S1].

***The effects of treatment are not consistent across treatment groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups [Figure S1]. It is likely that high inter-individual variation and the patchy distribution of OTUs across individuals limited the statistical power of the test. To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. Because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that that models would overfit the data. Despite this restriction, the models performed well (adenoma AUC=`r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC=`r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC=`r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. These results support the earlier community-wide analysis where we observed that the treatment had an impact on the overall community structure; however, the effect of treatment was not consistent across patients and diagnosis groups.

***Post-treatment samples from patients with carcinoma change towards a microbiota associated with normal***
Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons. To test this, we used an expanded cohort of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma [Table 2]. We then constructed Random Forest models to classify the pre and post-treatment samples as having their original diagnosis or having a normal colon. The models performed well (adenoma AUC=`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC=`r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC=`r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2). The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A]. Among the OTUs that were shared across the three models were those populations commonly considered as commensals (e.g. *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B]. Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [Figure S3]. Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model. A decrease in the positive probability would indicate that the colon more closely resembled that of a patient with a normal colon. There was no significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 4]. The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment. In fact, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) that were initially diagnosed with a carcinoma had a higher positive probability after treatment. One of those was re-diagnosed with carcinoma on the follow up visit. These results indicate that although there are changes in the microbiota associated with treatment, those experienced by the patients in the carcinoma group were the only ones where the change was directed towards what would be found in a normal colon

***Difficult to identify effects of specific treatments on the change in the microbiota.***
The type of treatment that the patients received varied across diagnosis groups. Those with adenomas and advanced adenomas received surgical resection (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "Y")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "Y")))`) or polyp removal during colonoscopy (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "N")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "N")))`) and those with carcinomas received surgical resection (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "no")))`), surgical resection with chemotherapy (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "no")))`), and surgical resection with chemotherapy and radiation (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "yes")))`). We focused on the patients with carcinoma and pooled those patients that received chemotherapy with those that received chemotherapy and radiation to improve our statistical power. We did not observe a significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P= `r chemo_rad_summary["dist", "bh"]`). Finally, the change in the positive probability was not significantly different between the two treatment groups (P=`r chemo_rad_summary["red_crc", "bh"]`). Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota

@@

Understanding the effect of treatment on the microbiota of those initially diagnosed with carcinomas may have important implications into understanding disease recurrence. The microbiota changes associated with treatment among patients initially diagnosed with carcinomas resulted in community structures that resembled patients with normal colons. This would suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at removing bacteria associated with them. Interestingly, 6 of the 26 patients diagnosed with carcinomas had an elevated probability of having carcinomas after treatment. This leads us to hypothesize that these individuals may be at a high risk of recurrence. Interestingly, the `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30% [@hellinger_reoperation_2006; @ryuk_predictive_2014]). Unfortunately, because of the design of our study, it was not possible to determine whether individuals who had an increased positive probability for carcinoma experienced recurrence. Regardless, it is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions, but to also assess the risk of developing them.

@@

***16S rRNA Gene Sequencing.***
Sequencing was completed as described by Kozich, et al. [@kozich_development_2013]. DNA extraction used the 96-well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf). The V4 variable region was amplified and the resulting product was split between three sequencing runs with normal, adenoma, and carcinoma evenly represented on each run. Each group was randomly assigned to avoid biases based on sample collection location. The pre and post-treatment samples were sequenced on the same run.

@@

***Statistical Analysis.***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis. Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`). Comparisons between probabilities as well as overall OTU differences between pre and post-treatment samples utilized a paired Wilcoxon ranked sum test. Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant. Unless otherwise stated the P-values reported are those that were BH corrected. Model rank importance was determined by obtaining the median MDA from the 100, 20 repeated 10-fold cross validation and then ranking from largest to smallest MDA.

@@

**Figure 1: General Differences between Adenoma, Advanced Adenoma, and Carcinoma Groups After Treatment.** A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`). Solid black points represent the median value for each diagnosis group. B) NMDS of the pre and post-treatment samples for the adenoma group. C) NMDS of the pre and post-treatment samples for the advanced adenoma group. D) NMDS of the pre and post-treatment samples for the carcinoma group.

@@

**Table 1: Demographic Data of pre and Post Treatment Cohort**
